### 2021-09-14 thesis/tex/oscillations.tex
update

\section{Neutrino Mixing}\label{ch:oscillation}
Japan 1998. The Super-Kamiokande experiment detects a $\nm$ deficit that could not be explained by any other mechanism
than the conversion of muon neutrinos to electron neutrinos. This was the first conclusive evidence of a completely new physics phenomenon: neutrino flavor
oscillations. An implication of the observation of oscillations is neutrino mass, 
which is identically zero acording to the Standard Model. Thus, we are required to extend the Standard Model to incorporate neutrino 
masses before we discuss the neutrino oscillations themselves.

@@

We consider a right-handed neutrino field, $\nu_R$. Since the electroweak gauge group $\text{SU}(2)_L \times U(1)_Y$ only couple to 
left-handed particles and right-handed antiparticles, $\nu_R$ transforms as a singlet under the Standard Model symmetry 
group $\mathrm{SU}(3)_{\mathrm{C}} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}$. 
This neutrino field is \emph{sterile} since it doesn't participate any of the Standard Model interactions.

We extend the Standard Model by adding a right-handed component of this field with neutrino Yukawa couplings $Y_{\alpha \beta}^{\prime \nu}$ to the Higgs-lepton Yukawa Lagrangian 
\begin{align}
    \mathcal{L}_{H}=-\left( \frac{v + H}{\sqrt{2}} \right) \left[\ell_{\alpha L}^{\prime} Y_{\alpha \beta}^{\prime \ell} \ell_{\beta R}^{\prime} + \nu_{\alpha L}^{\prime} Y_{\alpha \beta}^{\prime \nu} \nu_{\beta R}^{\prime}\right]\,.
\end{align}
Now, the Yukawa coupling matrices $Y^{\prime \ell}$ and $Y^{\prime \nu}$ are non-diagonal, with the former resulting in the charged lepton masses being non-definite. 
This can be remedied by diagonalizing $Y^{\prime \ell}$ with a unitary matrix $V^\ell$ as
\begin{align}
    V_{\alpha k L}^{\nu \dagger} Y^{\prime \nu}_{\alpha \beta} V_{\beta j R}^{\nu}=Y^{\nu}_{kj} \,.
\end{align}
Similarily, we diagonalize and $Y^{\prime \nu}$ with a unitary matrix $V^\nu$. We note that this will not yet generate neutrino masses, 
since no flavor rotation can make massless fields massive.
Now we state a crucial difference between the properties of the charged lepton and the neutrino fields.
While the charged lepton flavor eigenstate was uniquely determined by its mass eigenstate, the neutrino flavor is a superposition of mass eigenstates. 
This is because neutrinos are indirectly detected via the observation of its associated charged lepton, so there is no requirement of neutrino flavor eigenstates to have a definite mass. 
The flavor of a neutrino is then, by definition, the flavor of the associated charged lepton. 
This fact is denoted as giving the mass eigenstates Latin numerals and letters, while the flavor eigenstates stay as Greek letters.

@@

What form does this new matrix $U$ take? By construction, it provides the unitary transformation between the flavor and mass bases.
Any unitary $3\times3$ matrix can be parametrized using three mixing angles and six phases. However, not all phases affect the 
charged and weak currents and are thus not observable by us. Moreover, both the Lagrangian and the currents are invariant under global $U(1)$ transformations,
leaving only one physical phase for us to measure. We are now down to four degrees of freedom in the three dimensional case: three mixing angles
of the form $\sin{(\theta_{ij})}/\cos{(\theta_{ij})} = s_{ij}/c_{ij}$, and one phase of the form $e^{i\delta}$. The mixing angle $\theta_{ij}$ specifies 
the amount of rotation in the $i-j$ plane, allowing an $i$ eigenstate to mix with a $j$ eigenstate. The physical interpretation of
a mixing angle is then the degree of which the $\nu_i$ eigenstate mixes with the $\nu_j$ eigenstate. A non-zero mixing angle will then 
cause the flavor eigenstate $\nu_\alpha$ to be a superposition of the mass eigenstates that we allowed to mix. %TODO: nua = nu1 + nu2

@@

In this work, we will be using the best-fit values from~\cite{nufit} with the exception of the CP-violating phase $\delta_\text{CP}$. The $90\% $ CL values are
\begin{align}\label{eq:nufitparams}
    \theta_{12} = \SI{33.44}{\degree},\hspace{1em} \theta_{13} = \SI{8.57}{\degree},\hspace{1em} \theta_{23} = \SI{49.2}{\degree}, \hspace{1em} \delta_\text{CP} = 0\,.
\end{align}
Our neglect of the issue of CP invariance leaves the PMNS matrix real, with numerical values 
\begin{align}\label{eq:Uvalues}
    U = \begin{pmatrix}
        U_{e 1} & U_{e2} & U_{e3} \\
        U_{\mu 1} & U_{\mu 2} & U_{\mu 3} \\
        U_{\tau 1} & U_{\tau 2} & U_{\tau 3}
    \end{pmatrix} 
    = \begin{pmatrix}
        0.825 & 0.545 & 0.149 \\
        -0.455 & 0.485 & 0.746 \\
        0.334 & -0.684 & 0.649
    \end{pmatrix} \,.
\end{align}
In our study, we will only use neutrinos originating from conventional cosmic ray interactions in the atmosphere. 
In this so-called atmospheric neutrino flux, $\nm$ are the most abundant, while the $\nt$ flux is neglible. From Eq.~\eqref{eq:Uvalues}, w
e see that $U_{\mu\tau}$ is the largest $U_{\mu\beta}$ term,
suggesting that $\nm$ primarily mixes into $\nt$, at least in vacuum. It turns out that matter effects indeed preserves this ordering, making the $\nm \to \nt$ transition
the most abundant. This means we will be able to stringently constrain parameters relating to $\mu\tau$ mixing.

@@

We assume the neutrino masses $m_k$ to be extremely small compared to their associated energies $E_k$. 
Thus, $v\sim 1$, and $\abs{\vec{p}} \sim E$ making the energy-dispersion relation of Eq.~\ref{eq:disp} to first order:
\begin{align}\label{eq:ultra_rel}
    E_k &= \sqrt{\vec{p}^2 + m_k^2} \nonumber \\
        &= \vec{p}^2\sqrt{1 + \frac{m_k^2}{\vec{p}^2}} \nonumber \\
        &\approx E + \frac{m_k^2}{2E}
\end{align}
Hence, the exponential can be simplified, and simplifying the notation $P_{\na \to \nb}(t) \to P_{\ab}(t)$ we get 
\begin{align}
    P_{\ab}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)t/2E}\,.
\end{align}
Now, our approximation $v\approx 1$ implies $x\approx t $, thus
\begin{align}
    P_{\ab}(x) &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)x/2E} \nonumber \\
                       &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E})\,,
\end{align}
where we in the last step have defined the \emph{mass-squared difference} $\dm_{kj} = m_k^2 - m_j^2$. Since the oscillation probability depends on this quantity
rather than the individual masses, it is impossible to measure the mass $m_k$ through neutrino oscillations.
Squaring the unitarity condition $\sum_k U_{\a k}U_{\b k}^\dagger = \delta_{\a \b}$ yields 
\begin{align}
    \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 = \delta_{\a \b} - 2 \sum_{k>j}\Re[U_{\a k}^\dagger U_{\b k} U_{\a j} U_{\b j}^\dagger]
\end{align}
Since we take the PMNS matrix to be real and thus Hermitian, we drop the complex transpose in the following calculation. We have
\begin{align}\label{eq:Pab}%TODO: shorten?
    P_{\ab} &= \sum_{k,j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &=  \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 + \sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} +\sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} + 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\exp(-i\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            %&= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\cos(\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \sin^2{\left(\frac{\dm_{kj}x}{4E}\right)}\,,
\end{align}
which is the probability of neutrino vacuum oscillations.
The calculation for antineutrinos yield the same result if one continues to assume the realness of the mixing matrix. 
In Fig.~\ref{fig:vac_osc}, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a supressed $\nm \to \ne$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
\si{\GeV} oscillations are driven by the $\dm[31]$ mass-squared difference. Hence, the most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. The smallness of $U_{e3}$ then supresses the
amplitude of $\Pme$.
\begin{figure}
    \centering
    \includegraphics[width=0.9\textwidth]{figures/vac_osc.pdf}
    \caption{$\nm$ vacuum oscillations from Eq.~\ref{eq:Pab}}\label{fig:vac_osc}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=0.9\textwidth]{figures/w-boson.pdf} 
        \caption{Charged current weak interaction between an electron neutrino and an electron,
        mediated by either a $W^+$ or $W^-$ boson.}
    \end{subfigure}
    \quad
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=1\textwidth]{figures/z-boson.pdf} 
        \caption{Neutral current weak interaction between a neutrino of any flavor and a Earth-like fermion,
        mediated by a neutral $Z$ boson.}
    \end{subfigure}%TODO: they are different sizes
    \caption{Feymann diagrams showing the two interactions that neutrinos participate in according to the Standard Model.}\label{fig:w_and_z}
\end{figure}
In this work, we are concerned about the interactions with the neutrino and Earth-like matter, i.e.~electrons, protons, and neutrons. 
The possible interactions are shown in Fig.~\ref{fig:w_and_z}. The left panel shows that the only flavor that can undergo charged current (CC) 
interactions is the electron flavor. This is because the CC interaction requires a neutrino with flavor $\alpha$ to interact with its charged lepton 
partner $\alpha$. Since the Earth doesn't consist of any muons or tau particles, the $\nu_\mu$ and $\nu_\tau$ have no charged leptons to undergo CC interactions with. 
The right panel shows any neutrino flavor interaction via the neutral current (NC) with Earth-like matter\footnote{The Earth is entirely composed
of electrons, protons, and neutrons. Thus, the fundamental particles composing Earth are electrons, and up and down quarks. We refer to this as \emph{Earth-like matter}.}, mediated by the neutral $Z$ boson
The interaction mediated by the $W$ boson will give rise to a effective matter potential $V_{CC}$, while
the $Z$ boson is responsible for $V_{NC}$. Our task is now to find expressions for these.

We start with the effective Hamiltonian for the CC process. The Feynman rules for the left panel give us 
\begin{align}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, e \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, \ne \right]
\end{align}
By using the Fierz transformation 
\begin{align}
    \mathcal{L}^{V-A}(\psi_1,\psi_2,\psi_3,\psi_4) = \mathcal{L}^{V-A}(\psi_1,\psi_4,\psi_3,\psi_2)\,,
\end{align}
we can permute the terms inside the brackets, yielding
\begin{align}\label{eq:H_fierz}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, e \right]\,.
\end{align}
Now, lets consider a finite volume $V$ with electron states defined as 
\begin{align}\label{eq:e_states}
    \ket{e(p_e,h_e)} = \frac{1}{2E_eV}\, a_e^{(h_e)\dagger}(p_e)\ket{0}\,,
\end{align}
i.e. using the creation operator $a_e^{(h_e)\dagger}(p_e)$ to create electron states from vacuum with momenta $p_e$, energy $E_e$, and helicity $h_e$.
The density distribution of electrons in $V$ is $f(E_e,T)$, which we normalize to the total number of electrons as we integrate out the momenta $p_e$:
\begin{align}\label{eq:e_density}
    \int \dd p_e^3 f(E_e,T) = N_e V = n_e
\end{align}
Here, the electron density $N_e$ will ultimately determine the strength of the effective matter potential. 
To obtain the average effective Hamiltonian, project it on the electron states in Eq.~\ref{eq:e_states} and integrate over the density and sum over the helicities:

>>>>

\section{Neutrino Mixing}\label{ch:oscillation}
Japan 1998. The Super-Kamiokande experiment detects a $\nm$ deficit that could not be explained by any other mechanism
than the conversion of muon neutrinos to electron neutrinos. This was the first conclusive evidence of a completely new physics phenomenon: neutrino flavor
oscillations. An implication of the observation of oscillations is neutrino mass, 
which is identically zero according to the Standard Model. Thus, we are required to extend the Standard Model to incorporate neutrino 
masses before we discuss the neutrino oscillations themselves.

@@

We consider a right-handed neutrino field, $\nu_R$. Since the electroweak gauge group $\text{SU}(2)_L \times U(1)_Y$ only couple to 
left-handed particles and right-handed antiparticles, $\nu_R$ transforms as a singlet under the Standard Model symmetry 
group $\mathrm{SU}(3)_{\mathrm{C}} \times \mathrm{SU}(2)_{L} \times \mathrm{U}(1)_{Y}$. 
This neutrino field is \emph{sterile} since it doesn't participate in any of the Standard Model interactions.

We extend the Standard Model by adding a right-handed component of this field with neutrino Yukawa couplings $Y_{\alpha \beta}^{\prime \nu}$ to the Higgs-lepton Yukawa Lagrangian 
\begin{align}
    \mathcal{L}_{H}=-\left( \frac{v + H}{\sqrt{2}} \right) \left[\ell_{\alpha L}^{\prime} Y_{\alpha \beta}^{\prime \ell} \ell_{\beta R}^{\prime} + \nu_{\alpha L}^{\prime} Y_{\alpha \beta}^{\prime \nu} \nu_{\beta R}^{\prime}\right]\,.
\end{align}
Now, the Yukawa coupling matrices $Y^{\prime \ell}$ and $Y^{\prime \nu}$ are non-diagonal, with the former resulting in the charged lepton masses being non-definite. 
This can be remedied by diagonalizing $Y^{\prime \ell}$ with a unitary matrix $V^\ell$ as
\begin{align}
    V_{\alpha k L}^{\nu \dagger} Y^{\prime \nu}_{\alpha \beta} V_{\beta j R}^{\nu}=Y^{\nu}_{kj} \,.
\end{align}
Similarly, we diagonalize and $Y^{\prime \nu}$ with a unitary matrix $V^\nu$. We note that this will not yet generate neutrino masses, 
since no flavor rotation can make massless fields massive.
Now we state a crucial difference between the properties of the charged lepton and the neutrino fields.
While the charged lepton flavor eigenstate was uniquely determined by its mass eigenstate, the neutrino flavor is a superposition of mass eigenstates. 
This is because neutrinos are indirectly detected via the observation of its associated charged lepton, so there is no requirement for neutrino flavor eigenstates to have a definite mass. 
The flavor of a neutrino is then, by definition, the flavor of the associated charged lepton. 
This fact is denoted as giving the mass eigenstates Latin numerals and letters, while the flavor eigenstates stay as Greek letters.

@@

What form does this new matrix $U$ take? By construction, it provides the unitary transformation between the flavor and mass bases.
Any unitary $3\times3$ matrix can be parametrized using three mixing angles and six phases. However, not all phases affect the 
charged and weak currents and are thus not observable by us. Moreover, both the Lagrangian and the currents are invariant under global $U(1)$ transformations,
leaving only one physical phase for us to measure. We are now down to four degrees of freedom in the three dimensional case: three mixing angles
of the form $\sin{(\theta_{ij})}/\cos{(\theta_{ij})} = s_{ij}/c_{ij}$, and one phase of the form $e^{i\delta}$. The mixing angle $\theta_{ij}$ specifies the amount of rotation in the $i-j$ plane, allowing an $i$ eigenstate to mix with a $j$ eigenstate. The physical interpretation of
a mixing angle is then the degree to which the $\nu_i$ eigenstate mixes with the $\nu_j$ eigenstate. A non-zero mixing angle will then 
cause the flavor eigenstate $\nu_\alpha$ to be a superposition of the mass eigenstates that we allowed to mix. %TODO: nua = nu1 + nu2

@@

In this work, we will be using the best-fit values from~\cite{nufit} with the exception of the CP-violating phase $\delta_\text{CP}$. The $90\% $ CL values are
\begin{align}\label{eq:nufitparams}
    \theta_{12} = \SI{33.44}{\degree},\hspace{1em} \theta_{13} = \SI{8.57}{\degree},\hspace{1em} \theta_{23} = \SI{49.2}{\degree}, \hspace{1em} \delta_\text{CP} = 0\,.
\end{align}
Our neglect of the issue of CP invariance leaves the PMNS matrix real, with numerical values 
\begin{align}\label{eq:Uvalues}
    U = \begin{pmatrix}
        U_{e 1} & U_{e2} & U_{e3} \\
        U_{\mu 1} & U_{\mu 2} & U_{\mu 3} \\
        U_{\tau 1} & U_{\tau 2} & U_{\tau 3}
    \end{pmatrix} 
    = \begin{pmatrix}
        0.825 & 0.545 & 0.149 \\
        -0.455 & 0.485 & 0.746 \\
        0.334 & -0.684 & 0.649
    \end{pmatrix} \,.
\end{align}
In our study, we will only use neutrinos originating from conventional cosmic ray interactions in the atmosphere. 
In this so-called atmospheric neutrino flux, $\nm$ are the most abundant, while the $\nt$ flux is negligible. From Eq.~\eqref{eq:Uvalues}, w
e see that $U_{\mu\tau}$ is the largest $U_{\mu\beta}$ term,
suggesting that $\nm$ primarily mixes into $\nt$, at least in vacuum. It turns out that matter effects indeed preserves this ordering, making the $\nm \to \nt$ transition
the most abundant. This means we will be able to stringently constrain parameters relating to $\mu\tau$ mixing.

@@

We assume the neutrino masses $m_k$ to be extremely small compared to their associated energies $E_k$. 
Thus, $v\sim 1$, and $\abs{\vec{p}} \sim E$ making the energy-dispersion relation of Eq.~\ref{eq:disp} to first order:
\begin{align}\label{eq:ultra_rel}
    E_k &= \sqrt{\vec{p}^2 + m_k^2} \nonumber \\
        &= \vec{p}^2\sqrt{1 + \frac{m_k^2}{\vec{p}^2}} \nonumber \\
        &\approx E + \frac{m_k^2}{2E}
\end{align}
Hence, the exponential can be simplified, and simplifying the notation $P_{\na \to \nb}(t) \to P_{\ab}(t)$ we get 
\begin{align}
    P_{\ab}(t) = \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)t/2E}\,.
\end{align}
Now, our approximation $v\approx 1$ implies $x\approx t $, thus
\begin{align}
    P_{\ab}(x) &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} e^{-i(m_k^2-m_j^2)x/2E} \nonumber \\
                       &= \sum_{k,j} U^\dagger_{\a k}U_{\b k}U^\dagger_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E})\,,
\end{align}
where we in the last step have defined the \emph{mass-squared difference} $\dm_{kj} = m_k^2 - m_j^2$. Since the oscillation probability depends on this quantity
rather than the individual masses, it is impossible to measure the mass $m_k$ through neutrino oscillations.
Squaring the unitarity condition $\sum_k U_{\a k}U_{\b k}^\dagger = \delta_{\a \b}$ yields 
\begin{align}
    \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 = \delta_{\a \b} - 2 \sum_{k>j}\Re[U_{\a k}^\dagger U_{\b k} U_{\a j} U_{\b j}^\dagger]
\end{align}
Since we take the PMNS matrix to be real and thus Hermitian, we drop the complex transpose in the following calculation. We have
\begin{align}\label{eq:Pab}%TODO: shorten?
    P_{\ab} &= \sum_{k,j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &=  \sum_k \abs{U_{\a k}}^2\abs{U_{\b k}}^2 + \sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} +\sum_{k \neq j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2 \sum_{k>j}U_{\a k} U_{\b k} U_{\b j} U_{\a j} + 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \exp(-i\frac{\dm_{kj}x}{2E}) \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\exp(-i\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            %&= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \left[1-\cos(\frac{\dm_{kj}x}{2E})\right] \nonumber \\
            &= \delta_{\a \b} - 2\sum_{k > j} U_{\a k}U_{\b k}U_{\b j}U_{\a j} \sin^2{\left(\frac{\dm_{kj}x}{4E}\right)}\,,
\end{align}
which is the probability of neutrino vacuum oscillations.
The calculation for antineutrinos yield the same result if one continues to assume the realness of the mixing matrix. 
In Fig.~\ref{fig:vac_osc}, we show the energy spectra of these oscillations in the low \si{\GeV} range for neutrinos that travel \SI{12000}{\km}\footnote{This baseline
is approximately the Earth diameter, and will be the maximum travel distance which we will use later on.}.
In the left-most panel, we see a supressed $\nm \to \ne$ transition. We can understand this behavior by studying Eq.~\ref{eq:Pab} and the 
numerical values of the mixing matrix in Eq.~\ref{eq:Uvalues}.
\si{\GeV} oscillations are driven by the $\dm[31]$ mass-squared difference. Hence, the most important mixing matrix 
elements for $P_{\a\b}$ are $U_{\a3}, U_{\a1}$ and $U_{\b3}, U_{\b1}$. The smallness of $U_{e3}$ then supresses the
amplitude of $\Pme$.
\begin{figure}[!h]
    \begin{center}
    \includegraphics[width=0.9\textwidth]{figures/vac_osc.pdf}
    \caption{$\nm$ vacuum oscillations from Eq.~\ref{eq:Pab}}\label{fig:vac_osc}
    \end{center}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}{0.3\textwidth}
        \includegraphics[width=0.9\textwidth]{figures/w-boson.pdf} 
        \caption{Charged current weak interaction between an electron neutrino and an electron,
        mediated by either a $W^+$ or $W^-$ boson.}
    \end{subfigure}
    \quad
    \begin{subfigure}{0.3\textwidth}
        \vspace{1em}
        \includegraphics[width=1\textwidth]{figures/z-boson.pdf} 
        \caption{Neutral current weak interaction between a neutrino of any flavor and a Earth-like fermion,
        mediated by a neutral $Z$ boson.}
    \end{subfigure}
    \caption{Feymann diagrams showing the two interactions that neutrinos participate in according to the Standard Model.}\label{fig:w_and_z}
\end{figure}
In this work, we are concerned about the interactions with the neutrino and Earth-like matter, i.e.~electrons, protons, and neutrons. 
The possible interactions are shown in Fig.~\ref{fig:w_and_z}. The left panel shows that the only flavor that can undergo charged current (CC) 
interactions are the electron flavor. This is because the CC interaction requires a neutrino with flavor $\alpha$ to interact with its charged lepton 
partner $\alpha$. Since the Earth doesn't consist of any muons or tau particles, the $\nu_\mu$ and $\nu_\tau$ have no charged leptons to undergo CC interactions with. 
The right panel shows any neutrino flavor interaction via the neutral current (NC) with Earth-like matter\footnote{The Earth is entirely composed
of electrons, protons, and neutrons. Thus, the fundamental particles composing Earth are electrons, and up and down quarks. We refer to this as \emph{Earth-like matter}.}, mediated by the neutral $Z$ boson
The interaction mediated by the $W$ boson will give rise to a effective matter potential $V_{CC}$, while
the $Z$ boson is responsible for $V_{NC}$. Our task is now to find expressions for these.

We start with the effective Hamiltonian for the CC process. The Feynman rules for the left panel give us 
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, e \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, \ne \right]
By using the Fierz transformation 
    \mathcal{L}^{V-A}(\psi_1,\psi_2,\psi_3,\psi_4) = \mathcal{L}^{V-A}(\psi_1,\psi_4,\psi_3,\psi_2)\,,
we can permute the terms inside the brackets, yielding
\begin{align}\label{eq:H_fierz}
    H_{CC} = \frac{G_F}{\sqrt{2}}\left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right] \left[\bar e \,\gamma_\rho\, (1- \gamma^5)\, e \right]\,.
Now, lets consider a finite volume $V$ with electron states defined as 
\begin{align}\label{eq:e_states}
    \ket{e(p_e,h_e)} = \frac{1}{2E_eV}\, a_e^{(h_e)\dagger}(p_e)\ket{0}\,,
i.e. using the creation operator $a_e^{(h_e)\dagger}(p_e)$ to create electron states from vacuum with momenta $p_e$, energy $E_e$, and helicity $h_e$.
The density distribution of electrons in $V$ is $f(E_e,T)$, which we normalize to the total number of electrons as we integrate out the momenta $p_e$:
\begin{align}\label{eq:e_density}
    \int \dd p_e^3 f(E_e,T) = N_e V = n_e
Here, the electron density $N_e$ will ultimately determine the strength of the effective matter potential. 
To obtain the average effective Hamiltonian, project it on the electron states in Eq.~\ref{eq:e_states} and integrate over the density and sum over the helicities:
\begin{align}\label{eq:avg_H1}
    \avg{H_{CC}} &= \int \dd p_e^3\, \bra{e(p_e,h_e)} \times \frac{1}{2} \sum_{h_e} H_{CC}   f(E_e,T) \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \int \dd p_e^3 \bra{e(p_e,h_e)} \left[ \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \right]  f(E_e,T) \nonumber \\
           &\hspace{2cm}\times \frac{1}{2} \sum_{h_e} \left[\bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x) \right] \ket{e(p_e,h_e)} \nonumber \\
           &= \frac{G_F}{\sqrt{2}} \bar\ne \gamma^\rho\, (1- \gamma^5)\, \ne \int \dd p_e^3  f(E_e,T) \nonumber \\
           &\hspace{2cm} \times \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)}\,.
First, calculate the sum using trace technology
\begin{align}\label{eq:helicity_sum}
    \frac{1}{2} \sum_{h_e} \bra{e(p_e,h_e)} \bar e(x) \,\gamma_\rho\, (1- \gamma^5)\, e(x)   \ket{e(p_e,h_e)} &= \frac{1}{4E_e V} \sum_{h_e} \bar{u}_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, u_e^{h_e}(p_e) \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ \sum_{h_e} \bar{u}_e^{h_e}(p_e) u_e^{h_e}(p_e) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{1}{4 E_e V} \Trace{\left[ (\slashed{p}_e + m_e ) \,\gamma_\rho\, (1- \gamma^5)\, \right]} \nonumber \\
    &= \frac{(p_e)_\rho}{E_e V}\,.
Eq.~\ref{eq:avg_H1} now becomes 
\begin{align}\label{eq:avg_H2}
    \avg{H_{CC}} &= \frac{G_F}{\sqrt{2}E_e V} \bar\ne \, (1- \gamma^5)\, \ne \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T)\,.
Expand the integral, and use the fact that $\vec{p}_e$ is odd:
    \int \dd p_e^3\, \slashed{p}_e\, f(E_e,T) &= \int \dd p_e^3\, f(E_e,T) (\gamma^0 E_e - \vec{p}_e \cdot \vec{\gamma}) \nonumber \\
                                              &= \int \dd p_e^3\, f(E_e,T) \gamma^0 E_e \nonumber \\
                                              &= \gamma_0 E_e N_e V\,.
Inserting this into Eq.~\ref{eq:avg_H2}, we have
    \avg{H_{CC}} &= \frac{G_F N_e}{\sqrt{2}} \bar{\nu}_e \, (1- \gamma^5)\, \ne \gamma_0 \nonumber \\
            &= \sqrt{2}G_F N_e \bar{\nu}_{eE} \gamma^0 \nu_{eL}\,,
where the projection operator $(1- \gamma^5)$ in the first line ensures that only the left-hand compontent of the neutrino fields interact. Thus,
\begin{align}\label{eq:V_CC1}
    V_{CC} = \sqrt{2}G_F N_e\,, \quad H_{CC}\ket{\nu_k} = V_{CC}\ket{\nu_k}\,.
Here we see a crucial difference between the eigenvectors between the vacuum Hamiltonian defined in Eq.~\ref{eq:osc_1} and $H_{CC}$, namely
that the CC interactions happen in the flavor basis rather than in the mass basis\footnote{A similar calculation for the NC reveals the same fact:
neutrinos interact in the flavor basis.}. In other words,
neutrinos propagate in their mass eigenstates but interact in their flavor eigenstate. The mixing of mass eigenstates during propagation
determines if the flavor eigenstate has oscillated or not. The interactions still conserve lepton number since the interaction Lagrangian does,
but the kinetic Lagrangian will not. Hence, neutrinos violate lepton number convervation during propagation.
